---
title: "STAT 345 Midterm Project"
font: 12pt
date: "Due April 4"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

> "Our offense is like the Pythagorean Theorem. There is no answer!", *Shaquille O'Neal*

## The Background

Your role for the midterm project is that of data analyst intern at an NBA (professional basketball) team. Your direct supervisor (also part of the analytics team) has asked you to create a data visualization to illustrate how (or if) the team's shots have changed over time. After some initial clarifying questions, your supervisor confessed that they had seen some pretty cool shot charts at <http://savvastjortjoglou.com/nba-shot-sharts.html> and would like to extend the ideas a bit.

Your data for the midterm project may come from a variety of sources, including the NBA directly, as well as Basketball-Reference, HoopsHype, and others. There are several ways to access the data, but perhaps the simplest is through the `nbastatR` package, which serves as a wrapper (or set of helper functions) for these various data sources. There is a lot of potentially useful information that can be found using this package; unfortunately, the documentation is somewhat inconsistent. For example, the documentation at <https://www.rdocumentation.org/packages/nbastatR/versions/0.1.110202031> is pretty well-formatted and easy to navigate, but the current function names have changed a bit (see `?nbastatR` and click "Index"). (If you use the nbastatR package, you'll want to install the package using the following code. There are other packages with this type of data, though.) You are welcome to scrape other websites or use other packages if you choose.

```         
library(devtools)
devtools::install_github("abresler/nbastatR")
```

## The Tasks

1.  (30 points) Produce a graphic displaying the shot locations for a particular team over several years. Some notes:

    -   Colors should be chosen to reflect the team, if possible.

    -   There are likely many overlaid points -- handle this by either binning these by location, or use opacity.

    -   Incorporate information about whether the shot was made or not (shape, color, etc.).

    -   The graphic should be well-labeled, titled, etc.

    -   Start with a graph for a single year, then extend to several years. Up to 20 years of shot data is available. Either facet these by year or animate using the years.

    -   You'll want to figure out what the coordinates mean somehow. This might be through the documentation, but could also be determined using aspects of the data itself and the dimensions of an NBA court.

    -   Put a basketball court on the background of the image (you'll need to scale it appropriately).

    ```{r, force = TRUE}
    library(devtools)
            devtools::install_github("abresler/nbastatR@aba9179ef644f263387c1536d6ddd26104d79cf4")
    ```


```{r, fig.width=15, fig.height=8}
library(nbastatR) # Loaded this package so that I can get access to the nba data that I used in this code
library(ggplot2) # Loaded this package to be able to create the shot chart and adjust the appearance of it
library(magick) # I used this package to be able to read the basketball court image and have the ability to rotate it 
library(grid) # This package was added to make the overlay of data points and the basketball court image possible
library(gridExtra) # This package was added to be able to use the grid.arrange function to gather all the shot charts and put them together

seasons <- 2014:2018 # The seasons that I picked to analyze shot charts for the 76ers

Sys.setenv("VROOM_CONNECTION_SIZE" = 5000000) # Code was running just fine until the morning of 4/3, then I got an error about needing to increase the buffer size?

court_url <- "https://t3.ftcdn.net/jpg/01/88/00/92/360_F_188009274_RCkdTSELKU66ZBC3Mu28XsEpNgMez6fH.jpg" # The url of the basketball court image that I chose for the graphics
court_img_path <- "court.jpg" # The path of the image that I choose
court_img <- image_read(court_img_path) # This is added because I need to read the image path in order for it to be loaded into R for other manipulations to it
court_img_rotated <- image_rotate(court_img, 90) # The image first came out horizontally orientated and I wanted it vertical to align with the data, so I rotated it 90 degrees
court_img_raster <- as.raster(court_img_rotated) # I needed to add the raster function because it allows my image to be compatible with the ggplot2 package

court_dims <- c(-300, 300, -75, 900) # Dimensions of the basketball court

shot_graphic <- function(season) { # Function that is made to make the graphic with the basketball court and the data points
  shots_data <- teams_shots(teams = "Philadelphia 76ers", seasons = season, return_message = FALSE) # Retrieves data from nbastatR about the team shots for the 76ers for the seasons that I wanted. I added the return_message = FALSE because I did not want the extra, not important output that it gave me.
  
  ggplot(shots_data, aes(x = locationX, y = locationY)) +   
    annotation_custom(rasterGrob(court_img_raster, width = unit(1, "npc"), height = unit(1, "npc")), # annotation_custom is needed to add the image of the basketball court to the graphics, the "npc" argument is to make sure that the image fills up the plots
    xmin = court_dims[1], xmax = court_dims[2], # These two lines of code are written to set the boundaries for the image of the court which were previously stated above
    ymin = court_dims[3], ymax = court_dims[4]) +
    geom_point(aes(color = isShotMade, alpha = isShotMade), size = 1.2, alpha = 0.6) +  # Used to plot the data as points on the court, color of the points is determined through if the shot is made or not, alpha is responsible for changing the opacity of the points
    scale_color_manual(name = "Shot Result", values = c("FALSE" = "red", "TRUE" = "blue"), labels = c("Missed", "Made")) +  # This line of code was written to give the points certain colors (red or blue) based on of the shot was made or missed. I also changed the title and category names of the legend so that they read "Made" and "Missed" rather than "TRUE" and "FALSE"
    guides(color = guide_legend(override.aes = list(size = 4))) + # I needed to add this so that the dots that appeared in the legend were bigger than what they are in the plot
    theme_minimal() + # This was added to get rid of the grey background in the plots and make a more overall clean appearance
    labs(x = "Baseline", y = "Sideline", title = paste("76ers Shot Chart (", season - 1, "-", season, ")", sep = "")) + # This was added so that I could change the names of the x and y axis and give each of the plots different titles based on what the season is. The season - 1 part will start from 2018 and reduce by one for each graphic until it reaches 2014
    theme(legend.position = "top", # This is here to make sure that the legend is positioned on top of each of the graphics
          plot.title = element_text(hjust = 0.5), # I added this to center the title for each of the plots
          axis.text.x = element_blank(), # I wanted to get rid of the number values that displayed on the x axis
          axis.text.y = element_blank(), # Same as the x-axis, I wanted to get rid of the numbers that appeared on the y-axis
          panel.grid.major = element_blank(), # The next two lines of code were added to make the grid lines that were originally there go away for a cleaner look
          panel.grid.minor = element_blank()) +
    coord_fixed() + # I put this into the ggplot code because I did not want my output to be distorted at all
    ylim(-75, 857) # I set a limit on the dimension of the y-axis on the plot so that the points would all fit nicely
}

shot_charts <- lapply(seasons, shot_graphic) # lapply() is used here so that I can apply the shot graphic to all the seasons that I wanted to use it on

do.call(grid.arrange, c(shot_charts, ncol = 5)) # The do.call function is used here because I needed to display a list of many plots. The argument grid.arrange arranges the layout of the plots and the other argument is used to display the shot charts in 5 columns

# Results: After running the code, it will produce five different shot charts for the Philadelphia 76ers from the seasons 2013-2018. The legend of the charts shows that the made shots are blue and the missed shots are red. In summary, we can see how the shot selection and shots made/missed percentages have changed over the course of five seasons.

```

2.  (30 points) Summarize the graphic/series of graphics into a digestible, bullet-point brief report for front-office staff. Some notes:

    -   The main body of the report should be very brief -- just the graphic(s) and the bullet-pointed list of findings, which should be short and clear.
    -   Include a more detailed explanation of these bullet points, for further reading by those interested. This section should follow the bullet-point section, but should be organized similarly for reference.
    -   Your report to the front-office shouldn't include any code.
    -   This report should be generated using RMarkdown. However, the choice of output type (Word, PDF, or HTML) is up to you (you could even make slides if you want to).

3.  (30 points) Write and document clean, efficient, reproducible code. Some notes:

    -   This code will be viewed by your direct supervisor.
    -   The code file should include your code to gather, join, and clean the data; the code to generate the graphic(s) presented; and your commentary on the results (so, a single .rmd file, or an .rmd file that sources an .r file).
    -   Your code should be clean, organized, and reproducible. Remove unnecessary/scratch/exploratory code.
    -   Your code should be well commented. In particular, any decisions or judgement calls made in the analysis process should be explained/justified. Sections of code should be identified even if not functionalized (including purpose, data/argument inputs, analysis outputs).

4.  (10 points) Above and Beyond. Choose either option below. You are welcome to explore both, but only one is required.

-   Option 1: Explore the data a bit, and create a graphic that uses (or incorporates) different information than what was used above. Some notes:
    -   Create an additional graphic that incorporates at least one additional variable not previously used (this should add to the graphic in part 1). The additional data should be drawn from a different dataset (function call) than the original graphic used. These two (or more) datasets may need to be joined appropriately.
    -   You can either add more information to the plot above, or create a different plot.
    -   Formatting, labelling, etc. are all important here too.
    -   Adding marginal densities or other "bells and whistles" might offer additional insight.
    -   This graphic should be included at the end of the report (after the more detailed explanations).
    -   You should include a brief description of the graphic (highlighting the different/additional information used).
-   Option 2: If the NBA were to incorporate a 4-point shot, where would you draw a 4-point arc? Some notes:
    -   You likely should base your decision at least partly on proportion of shots made from a given range. You might consider an expected value calculation here.
    -   Your arc can be shaped as you see fit; simple arcs are sufficient for this exploration.
    -   Provide an example of a consequence (positive or negative) if a 4-point shot was incorporated. (e.g., "my_favorite_player's season point total would increase by x%")
    -   You do not need to create a plot representing your arc, though you are welcome to do so!

```{r}
head(shots_data)
```


```{r, warning = FALSE}
# Above and Beyond Option 1
library(dplyr) # Needed to add this package in order to use the summarize function, although it was working without it yesterday

shots_data <- teams_shots(teams = "Philadelphia 76ers", seasons = 2018, return_message = FALSE) # Retrieves data from nbastatR about the team shots for the 76ers for the season that I wanted. I added the return_message = FALSE because I did not want the extra, not important output that it gave me.

player_shots <- shots_data %>%
  filter(namePlayer == "Joel Embiid") %>% # This will filter out only the shots that Joel Embiid took, which is what I want
  mutate(shot_distance = sqrt(locationX^2 + locationY^2)) %>% # The calculation used to find the relative distance from where the shots were taken (not exact). This is called the Euclidean distance, in which I had to do some research about and found that it is as simple as the distance of the line segment between two points
  group_by(namePlayer, locationX, locationY) %>% # This was needed to make sure that I am getting the shots of Embiid and the locations of the shots
  summarize(shot_count = n(), avg_shot_distance = mean(shot_distance),
            .groups = "drop") # This part of the code takes into account the amount of shots that were taken at each location and will also do the calculation of the average length of the shot because some points on the plot have multiple shots taken from that same point. I did not have the .groups = "drop" argument originally, but I needed to add it because I got the error, "`summarise()` has grouped output by 'namePlayer', 'locationX'. You can override using the `.groups` argument."

court_url <- "https://t3.ftcdn.net/jpg/01/88/00/92/360_F_188009274_RCkdTSELKU66ZBC3Mu28XsEpNgMez6fH.jpg" # The url of the basketball court image that I choose for the graphic
court_img_path <- "court.jpg" # The path of the image that I choose
court_img <- image_read(court_img_path) # This is added because I need to read the image path in order for it to be loaded into R for other manipulations to it
court_img_rotated <- image_rotate(court_img, 90) # Image first came out horizontally orientated and I wanted it vertical to align with the data, so I rotated it 90 degrees
court_img_raster <- as.raster(court_img_rotated) # I needed to add the raster function because it allows my image to be compatible with the ggplot2 package

ggplot(player_shots, aes(x = locationX, y = locationY)) +
  annotation_custom(rasterGrob(court_img_rotated, width = unit(1, "npc"), height = unit(1, "npc")), # annotation_custom is needed to add the image of the basketball court to the graphic, the "npc" argument is to make sure that the image fills up the plot
  xmin = -300, xmax = 285, ymin = -75, ymax = 900) + # These are the dimensions for both axis of the graphic
  geom_point(aes(color = avg_shot_distance, size = shot_count), alpha = 0.5) +  # This was added to get the points to show up on the plot. Along with that, I needed to add arguments to have the color of the points be related to how far away the shot was and have the size of the points be related to how many times that shot was taken
  scale_color_gradientn( # This function was needed in order for me to create the colors that I wanted the points to be
    colors = c("green", "yellow", "red"), # The three colors that I chose for distance levels
    breaks = c(60,160,260), # This is added for the spacing of the labels in the legend
    labels = c("Short", "Mid-range", "Long") # The names of the labels that I put in the legend
  ) + 
  scale_size(range = c(1, 7)) + # This was put in here so that I am able to adjust the size of the smallest point all the up to the biggest point displayed in the graphic
  coord_fixed() + # I put this into the ggplot code because I did not want my output to be distorted at all
  theme_minimal() + # This was added to get rid of the grey background in the plots and make a more overall clean appearance
  labs(title = "Shot Distance Chart for Joel Embiid", # This is the title of the graphic that I made
       subtitle = "2017-18 Season", # This is the subtitle of the graphic
       x = "Baseline", # The title of the x axis
       y = "Sideline", # The title of the y-axis
       color = "Shot Distance", # The title of the color of the points legend
       size = "Shots Taken") + # The title of the size of the points legend
  theme(
    plot.title = element_text(hjust = 0.5), # I added this line and the next to center the title and subtitle for the graphic
    plot.subtitle = element_text(hjust = 0.5), 
          axis.text.x = element_blank(), # I wanted to get rid of the number values that displayed on the x axis
          axis.text.y = element_blank(), # Same as the x-axis, I wanted to get rid of the numbers that appeared on the y-axis
          panel.grid.major = element_blank(), # The next two lines of code were added to make the grid lines that were originally there go away for a cleaner look
          panel.grid.minor = element_blank()) +
  ylim(-75, 857) # I set a limit on the dimension of the y-axis on the plot so that the points would all fit nicely

```

## The Deliverables

1.  Upload your report and code file(s) to GitHub by 11:00am on Friday, April 4.
2.  Submit (on Canvas) your report, code, and link to your GitHub repository by 11:00am on Friday, April 4.
